---
title: "Machine Learning Prediction Project"
author: "Lisa Anderson"
date: "February 23, 2018"
output: html_document
---

##Introduction   
Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively.   
These type of devices are part of the quantified self movement - a group of enthusiasts who take measurements about themselves regularly to improve their health,   
to find patterns in their behavior, or because they are tech geeks.   
One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it.
The goal of this project is to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants to predict the way in which they performed (the "classe").
They were asked to perform barbell lifts correctly and incorrectly in 5 different ways.   
   
More information is available at <http://web.archive.org/web/20161224072740/http:/groupware.les.inf.puc-rio.br/har> (see the section on the Weight Lifting Exercise Dataset).   
   

##Cleaning and Preparing Data   
```{r}
library(caret)
set.seed(123)
trainUrl <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"   
testUrl <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"   
training <- read.csv(url(trainUrl), na.strings=c("NA","#DIV/0!",""))   
testing <- read.csv(url(testUrl), na.strings=c("NA","#DIV/0!",""))   
```  
   
Look at summary of the data. There are 160 variables, so we will need to narrow down this list by choosing predicators.   
```{r}
str(training)   
```     
Split training data further into training and test sets to save the original test data as a validation set.   
```{r}   
inTrain <- createDataPartition(training$classe, p=0.7, list=FALSE)  
train1<- training[inTrain, ]   
train.test <- training[-inTrain, ]   
```      
   
Remove Near Zero Variance variables because there are non-informative.   
This reduces the number of variables to 128 from 160 in the training data.   
```{r}    
nzvCol <- nearZeroVar(train1)   
train1 <- train1[,-nzvCol]   
train.test <- train.test[,-nzvCol]   
```   
   
There are still many columns that contain mostly NAs. We now delete columns of the training set that contain over 95% NAs    
and remove the same column names from the test data.   
This has further reduced the number of variables to 59.   ##Building the Model      

```{r}      
train1 <- train1[, colSums(is.na(train1))/(nrow(train1)) <=.95]   
train.test <-train.test[, colnames(train1)]   
```   
   
Lastly, we need to remove the columns that shouldn't be used as predictors (X, user_name, raw_timestamp_part_1, raw_timestamp_part_2, cvtd_timestamp).   
The names of the subjects and the timestamps would not be useful variables for the prediction model. This reduces the number of columns to 54.   
```{r} 
train1 <- train1[, -(1:5)]   
train.test <- train.test[, -(1:5)]   
```   
   
    
I selected random forest model and then fit this model on the train1 data.   
Controlling the tuning parameters by choosing a 3-fold cross validation method (method="cv" , number=3) will reduce the runtime of the model.   
If you don't choose parameters, the model may take too long to run.   
   
```{r} 
fitControl <- trainControl(method="cv", number=3, verboseIter=F)   
modfit <- train(classe ~ ., data=train1, method="rf",trControl=fitControl)   
```   
Viewing the final model tells you how many trees were used and the amount of variables at each split.   
```{r}
modfit$finalModel   
```   
   
## Evaluating Model Accuracy   
     
I used the model to predict the classe in the train.test data set.   
```{r}
pred1 <- predict(modfit, newdata=train.test)   
```    
   
The confusion matrix shows predictions vs the actuals for the classe column and has determined the accuracy of the model to be 99.78%.      
```{r}
confusionMatrix(train.test$classe, pred1)   
```     
Since this model produced results with 99.78% accuracy, I will not be trying any other models.   
The predicted out of sample error rate is .22%.   
   
## Use Selected Model to Predict on Test Set      
   
Now that the model has been chosen and tested, we will go back to the original training set, clean up the data,   
and fit the model as we did in the above steps with the further split training sets.   
   
Remove near zero variance variables.   
```{r}
nzvCol <- nearZeroVar(training)   
training <- training[,-nzvCol]   
testing <- testing[,-nzvCol]   
```   
   
Remove columns that are mostly NAs.   
In order to match the columns in testing set to training set, I created created a set of the training data without the classe column (column 59).   
Then I removed the columns from testing that did not match.   
```{r}
training <- training[, colSums(is.na(training))/(nrow(training)) <=.95]   
training.noclasse <- training[,-59]   
testing <- testing[,colnames(training.noclasse)]   
```   
    
Remove first 5 columns because they are not necessary.   
```{r}
training <- training[, -(1:5)]   
testing <- testing[, -(1:5)]   
```   
   
Fit the model to the original training data.   
```{r}
fitControl <- trainControl(method="cv", number=3, verboseIter=F)   
modfit1 <- train(classe ~ ., data=training, method="rf", trControl=fitControl)   
```   
   
##Predict on Test Set   
   
Predict the classe column for the testing dataset and print the results.   
```{r}
pred2 <- predict(modfit1, newdata=testing)   
pred2   
```     
  
##Conculusion   
   
Above are the predictionsfor the classe column in the testing data set using the random forest model that I built,   
which was trained on the training data set.   

